Determinism, Complexity, and Predictability in Computer Performance

### 2014-01-10 pecomplexity.tex
update some about word length

In practice, calculating permutation entropy involves choosing a good value for
the wordlength $n$. The key consideration here is that the value be large enough
that forbidden ordinals are discovered, yet small enough that reasonable
statistics over the ordinals are gathered: e.g., $\displaystyle n =
\operatornamewithlimits{argmax}_\ell\{T > 100 \ell!\}$, assuming an average of
100 counts per ordinal. In the literature, $3\le n \le 6$ is a standard choice
---generally without any formal justification. In theory, the permutation
entropy should reach an asymptote with increasing $n$, but that requires an
arbitrarily long time series. In practice, what one should do is calculate the
\emph{persistent} permutation entropy by increasing $n$ until the result
converges, but data length issues can intrude before that convergence is
reached.

Table~\ref{tab:pe} shows the permutation entropy results for the
examples considered in this paper, with the nRMSPE prediction
accuracies from the previous section included alongside for easy
comparison.

@@

  \begin{table}[htbp]
   \caption{Prediction error (in nRMSPE) and permutation entropy (for
     different wordlengths $n$}
   \label{tab:pe}
   \centering
   \begin{tabular}{|c|c|c|c|c|c|}
     \hline
      {\bf cache misses} & error   &$n=4$&$n=5$&$n=6$\\
     \hline
     \tt{row\_major}  & 0.0324  &0.6751&0.5458&0.4491\\
     \hline
     \tt{col\_major}  & 0.0080  &0.5029&0.4515&0.3955 \\
     \hline
     \tt{403.gcc}  & 0.1416  &0.9916&0.9880&0.9835\\
     \hline
     \tt{482.sphinx3}  & 0.2032  &0.9913&0.9866&0.9802 \\
      \hline
      \hline
      {\bf insts per cyc} & error   &$n=4$&$n=5$&$n=6$\\
     \hline
      \tt{row\_major} & 0.0778  &0.9723&0.9354&0.8876\\
     \hline
      \tt{col\_major} & 0.0161   &0.8356&0.7601&0.6880\\
     \hline
     \tt{403.gcc} & 0.2033   &0.9862&0.9814&0.9764\\
     \hline
     \tt{482.sphinx3} & 0.3670 & 0.9951&0.9914&0.9849 \\
      \hline
   \end{tabular}
 \end{table}
The relationship between prediction accuracy and the permutation
entropy (PE) is as we conjectured: performance traces with high
PE---those whose temporal complexity is high, in the sense that little
information is being propagated forward in time---are indeed harder to
predict using the simple deterministic forecast model described in the
previous section.  The effects of changing $n$ are also interesting:
using a longer wordlength generally lowers the PE---a natural
consequence of finite-length data---but the falloff is less rapid in
some traces than in others, suggesting that those values are closer to
the theoretical asymptote that exists for perfect data.  The
persistent PE values of 0.5--0.6 for the {\tt row\_major} and {\tt
  col\_major} cache-miss traces are consistent with dynamical chaos,
further corroborating the results of~\cite{mytkowicz09}.  (PE values
above 0.97 are consistent with white noise.)  Interestingly, the
processor-load traces for these two microkernels exhibit more temporal
complexity than the cache-miss traces.  This may be a consequence of
the lower baseline value of this time series.

>>>>

In practice, calculating permutation entropy and weighted permutation entropy
involves choosing a good value for the word length $n$. The primary
consideration is that the value be large enough that forbidden ordinals are
discovered, yet small enough that reasonable statistics over the ordinals are
gathered: e.g.:
\begin{align*}
n = \operatornamewithlimits{argmax}_\ell \{ T \gtrapprox 100 \ell! \},
\end{align*}
assuming an average of 100 counts per ordinal is sufficient. In the literature,
$3 \le n \le 6$ is a standard choice --- generally without any formal
justification. In theory, the permutation entropy should reach an asymptote with
increasing $n$, but that requires an arbitrarily long time series. In practice,
what one should do is calculate the \emph{persistent} permutation entropy by
increasing $n$ until the result converges, but data length issues can intrude
before that convergence is reached.

% Table~\ref{tab:pe} shows the permutation entropy results for the
% examples considered in this paper, with the nRMSPE prediction
% accuracies from the previous section included alongside for easy
% comparison.

@@

 %   \caption{Prediction error (in nRMSPE) and permutation entropy (for
 %     different wordlengths $n$}
 %   \label{tab:pe}
 %      {\bf cache misses} & error   &$n=4$&$n=5$&$n=6$\\
 %     \tt{row\_major}  & 0.0324  &0.6751&0.5458&0.4491\\
 %     \tt{col\_major}  & 0.0080  &0.5029&0.4515&0.3955 \\
 %     \tt{403.gcc}  & 0.1416  &0.9916&0.9880&0.9835\\
 %     \tt{482.sphinx3}  & 0.2032  &0.9913&0.9866&0.9802 \\
 %      {\bf insts per cyc} & error   &$n=4$&$n=5$&$n=6$\\
 %      \tt{row\_major} & 0.0778  &0.9723&0.9354&0.8876\\
 %      \tt{col\_major} & 0.0161   &0.8356&0.7601&0.6880\\
 %     \tt{403.gcc} & 0.2033   &0.9862&0.9814&0.9764\\
 %     \tt{482.sphinx3} & 0.3670 & 0.9951&0.9914&0.9849 \\

The relationship between prediction accuracy and the permutation
entropy (PE) is as we conjectured: performance traces with high
PE---those whose temporal complexity is high, in the sense that little
information is being propagated forward in time---are indeed harder to
predict using the simple deterministic forecast model described in the
previous section.  The effects of changing $n$ are also interesting:
using a longer wordlength generally lowers the PE---a natural
consequence of finite-length data---but the falloff is less rapid in
some traces than in others, suggesting that those values are closer to
the theoretical asymptote that exists for perfect data.  The
persistent PE values of 0.5--0.6 for the {\tt row\_major} and {\tt
  col\_major} cache-miss traces are consistent with dynamical chaos,
further corroborating the results of~\cite{mytkowicz09}.  (PE values
above 0.97 are consistent with white noise.)  Interestingly, the
processor-load traces for these two microkernels exhibit more temporal
complexity than the cache-miss traces.  This may be a consequence of
the lower baseline value of this time series.
